Transactional Memory Concurrency Verification with Landslide

### 2018-03-15 htm/paper.tex
buncha minor prose stuff

This motivates the use of stateless model checking (MC) \cite{verisoft}
to comprehensively verify these transactional programs,
fast paths failure paths and all.
MC aims to force the system to execute all possible thread interleavings under a given test case,
exhaustively checking for bugs or verifying their absence in the corresponding state space.
Many such model checkers exist, varying in
interleaving granularity, memory analysis, types of programs checked, and search ordering strategy
\cite{chess,dbug-ssv,spin,inspect,r4,portend,samc,mcr,randomized-scheduler}.
This work builds upon Landslide \cite{landslide,quicksand,landslide-phdthesis},
a simulator-based tester which checks both user- and kernel-level programs
and incorporates data-race analysis \cite{tsan,fasttrack} to find new preemption points
at memory access granularity.
Our contributions are as follows:

@@

MCs may instrument programs to introduce thread switches at varying granularity, which affects the number of operations $n$.
%The number of operations depends on how finely-grained the MC instruments the program to introduce thread switches.
Some target distributed systems, instrumenting only message-passing events \cite{modist};
some run multithreaded programs natively, instrumenting only the pthread API for performance \cite{dbug-ssv};
and some insert compiler instrumentation on statically-identified memory accesses \cite{chess,inspect}.
Landslide traces every memory access through the use of a simulated environment \cite{bochs},
which is important for identifying data races to use as new preemption points \cite{fasttrack,djit,quicksand},
as well as for identifying when a memory conflict may cause transaction aborts.
With regard to checking for bugs,
the ``model'' the name refers to being checked may be
an external formal specification,
the program's own internal consistency checks,
or a set of expected properties encoded in the tool itself.
Landslide uses the latter two cases, checking for assertion failures
as well as deadlocks, use-after-frees, and segfaults.
For this work we also detect use of {\tt xend} outside of a transaction as a bug.

@@

\begin{proof}
	We case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/write or write/write)
	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.
	If so, then the hardware will abort \tii's transaction, discarding the effects of $\tiiat{\beta_1}\dots\tiiat{\beta_n}$
	and jumping to $\tiiat{\phi_1}$,
	satisfying case 1.
	Otherwise, by DPOR's definition of transition dependence \cite{dpor,landslide-phdthesis},
	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
	may be successfully executed until transaction commit,
	and reordering them produces an equivalent interleaving,
	satisfying case 2.
\end{proof}

@@

Our implementation of HTM has been incorporated by the Landslide maintainers upstream.
The repository is available open-source at \url{https://github.com/bblum/landslide}.
Programs should be ported to the Pebbles userland \cite{thrlib,kspec},
their use of compiler HTM intrinsics should be replaced with the Landslide stubs provided in {\tt 410user/inc/htm.h},
and HTM nondeterminism can then be enabled with the {\tt -X} command-line flag.
We have also extended its Iterative Deepening implementation \cite{quicksand}
with an option ({\tt -M} flag) to
to optimize for completion time by prioritizing the maximal state space job and cancelling all others,
while maintaining its soundness guarantee,
which we will use in our evaluation.
All test cases therein are also available in the repository linked above.
%Several example tests are also provided:
%{\tt htm1} demonstrates the bug in Figure~\ref{fig:htm-example},
%and {\tt htm2} demonstrates Figure~\ref{fig:htm-fixed}'s fix.

@@

		{\bf New bugs.}
		In addition to the bugs we intentionally wrote in {\tt htm1} and {\tt swapbug},
		to our pleasant surprise
		Landslide also found a previously-unknown bug in the transactional AVL tree,
		exposed by {\tt avl\_insert} with any parameters higher than (2,1).
		Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
		% landslide-trace-1521093835.18927.html
		This manifested
		alternately
		%in (2,2) and (3,1)
		as a segfault
		% landslide-trace-1521092404.23772.html
		and
		%in (2,3)
		as a consistency-check assertion failure.
		The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
		in practice (between it and {\tt \_xbegin()}),
		whereas MC is blind to such matters of chance.%
		\footnote{Considering the loop does not affect the test's possible behaviours,
		only its {\em likely} ones,
		we removed it in Landslide's version of the test to keep the state space size manageable.}
		%We removed the loop in Landslide's version of the test
		%%To make this test Landslide-friendly, we removed that loop,
		%which reduces the state space size, but maintains the program's semantics.}
		Moreover, we conclude that even Figure~\ref{fig:htm-fixed}'s protocol's very proposer % こそさえ
		getting it wrong motivates the need for MC on such programs,
		and perhaps also means TSX primitives should be used only behind higher-level abstractions
		such as lock elision \cite{lock-elision}.

% TODO syntax hilighting
\begin{figure}[t]
	\begin{center}
		\begin{tabular}{l}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{while (\_retry);} \\
		\texttt{if (\_xbegin() == SUCCESS) \{} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_xend();} \\
		\texttt{\} else \{} \\
		\texttt{~~~~pthread\_mutex\_lock(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = true;} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_retry = false;} \\
		\texttt{~~~~pthread\_mutex\_unlock(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption{Unmodified code from {\tt htmavl.cpp} with the previously-unknown bug found by Landslide.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	%\item

@@

			& 3,1 & \ETAdag{8d 11h} & \ETAdag{411312}	\\
			%[JOB 8] progress: 103873/411312 brs (25.254021%), ETA 8d 11h 36m 0s (elapsed 59m 42s)
			%[JOB 8] TIMED OUT (25.254021%; ETA 8d 11h 36m 0s)
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{1083.35}	& \ints{40062}	\\
			%& 2,2 & \multicolumn{2}{c}{\em (bug!)} \\
			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{10d 16h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{877.44}& \ints{28635}	\\
			& 2,2 & \ETAdag{2d 6h} & \ETAdag{5925634} \\
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

@@

\begin{table}[t]
	\begin{center}
	\begin{tabular}{l|r|c}
		\bf Landslide mode & \bf total int's & \cpu{\bf cpu (s)} \\
		\hline
		{\sf verif} ({\tt -M})		& 403  & \cpu{~~14.95 $\pm$ 00.17} \\
		{\sf reinstrument}		& 403  & \cpu{~~24.72 $\pm$ 00.10}	\\
		{\sf quicksand} (no {\tt -M})	& 2128 & \cpu{120.53 $\pm$ 18.62}	\\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) in various modes.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants2}
\end{table}

>>>>

This motivates the use of stateless model checking (MC) \cite{verisoft}
to comprehensively verify these transactional programs,
fast paths failure paths and all.
MC aims to force the system to execute all possible thread interleavings under a given test case,
exhaustively checking for bugs or verifying their absence in the corresponding state space.
Many such model checkers exist, varying in
interleaving granularity, memory analysis, types of programs checked, and search ordering strategy
\cite{chess,dbug-ssv,spin,inspect,r4,portend,samc,mcr,randomized-scheduler}.
This work builds upon Landslide \cite{landslide}, %,quicksand,landslide-phdthesis},
a simulator-based tester which checks both user- and kernel-level programs
and incorporates data-race analysis \cite{tsan,fasttrack} to find new preemption points
at memory access granularity.
Our contributions are as follows:

@@

MCs may instrument programs to introduce thread switches at varying granularity, which affects the number of operations $n$.
%The number of operations depends on how finely-grained the MC instruments the program to introduce thread switches.
Some target distributed systems, instrumenting only message-passing events \cite{modist};
some run multithreaded programs natively, instrumenting only the pthread API for performance \cite{dbug-ssv};
and some insert compiler instrumentation on statically-identified memory accesses \cite{chess,inspect}.
Landslide traces every memory access through the use of a simulated environment \cite{bochs},
which is important for identifying data races to use as new preemption points \cite{quicksand},
as well as for identifying when a memory conflict may cause transaction aborts.
With regard to checking for bugs,
the ``model'' the name refers to being checked may be
an external formal specification,
the program's own internal consistency checks,
or a set of expected properties encoded in the tool itself.
Landslide uses the latter two cases, checking for assertion failures
as well as deadlocks, use-after-frees, and segfaults.
For this work we also detect use of {\tt xend} outside of a transaction as a bug.

@@

\begin{proof}
	We case on whether the operations by \tjj and/or \tkk have any memory conflicts (read/write or write/write)
	with $\tiiat{\beta_1}\dots\tiiat{\beta_n}$.
	If so, then the hardware will abort \tii's transaction, discarding the effects of $\tiiat{\beta_1}\dots\tiiat{\beta_n}$
	and jumping to $\tiiat{\phi_1}$,
	satisfying case 1.
	Otherwise, by DPOR's definition of transition dependence \cite{dpor}, %,landslide-phdthesis},
	$\tiiat{\beta_{b+1}}\dots\tiiat{\beta_n}$ is independent with the transitions of \tjj and \tkk,
	may be successfully executed until transaction commit,
	and reordering them produces an equivalent interleaving,
	satisfying case 2.
\end{proof}

@@

Our implementation of HTM-equivalent semantics has been incorporated by the Landslide maintainers upstream.
It is available open-source at \url{https://github.com/bblum/landslide}.
Programs should be ported to the Pebbles userland \cite{thrlib,kspec},
their use of compiler HTM intrinsics should be replaced with the Landslide stubs provided in {\tt 410user/inc/htm.h},
and HTM nondeterminism can then be enabled with the {\tt -X} command-line flag.
We have also extended its Iterative Deepening implementation \cite{quicksand}
with a new option ({\tt -M} flag) to
to optimize for completion time by prioritizing the maximal state space job and cancelling all others
(which maintains its soundness guarantee)
which we will use in our evaluation.
All test cases therein are also available in the repository linked above.
%Several example tests are also provided:
%{\tt htm1} demonstrates the bug in Figure~\ref{fig:htm-example},
%and {\tt htm2} demonstrates Figure~\ref{fig:htm-fixed}'s fix.

@@

		{\bf New bugs.}
		In addition to the bugs we intentionally wrote in {\tt htm1} and {\tt swapbug},
		to our pleasant surprise
		Landslide also found a previously-unknown bug in the transactional AVL tree,
		exposed by {\tt avl\_insert} with any parameters higher than (2,1).
		Figure~\ref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
		% landslide-trace-1521093835.18927.html
		This manifested
		alternately
		%in (2,2) and (3,1)
		as a segfault
		% landslide-trace-1521092404.23772.html
		and
		%in (2,3)
		as a consistency-check assertion failure.
		The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
		in practice (between it and {\tt \_xbegin()}),
		whereas MC is blind to such matters of chance.%
		% NB. In case a reader doesn't believe this footnote,
		% I put the loop back in (with thr_yield(-1) therein) to restore the original state space,
		% and found the bug with these statistics:
		% [JOB 28] progress: 1957/3005897 brs (0.065105%), ETA 6414673y 121d 18h 28m 59s (elapsed 3m 15s)
		% [JOB 28] BUG FOUND: landslide-trace-1521099044.76395.html (1958 interleavings tested; 3 preemptions; job time 3m 15s; pldi time 746832861; new-fixed pldi cputime 823243872)
		\footnote{Considering the loop does not affect the test's possible behaviours,
		only its {\em likely} ones,
		we removed it in Landslide's version of the test to keep the state space size manageable.}
		%We removed the loop in Landslide's version of the test
		%%To make this test Landslide-friendly, we removed that loop,
		%which reduces the state space size, but maintains the program's semantics.}
		Moreover, we conclude that even Figure~\ref{fig:htm-fixed}'s protocol's very proposer % こそさえ
		getting it wrong motivates the need for MC on such programs,
		and perhaps also means TSX primitives should be used only behind higher-level abstractions
		such as lock elision \cite{lock-elision}.

% TODO syntax hilighting
\begin{figure}[t]
	%\begin{center}
		\begin{tabular}{l}
		%\texttt{void count() \{} \\
		%\texttt{~~~~for (int i = 0; i < 1000; i++) \{} \\
		\texttt{while (\_retry);} \\
		\texttt{if (\_xbegin() == SUCCESS) \{} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_xend();} \\
		\texttt{\} else \{} \\
		\texttt{~~~~pthread\_mutex\_lock(\&\_tree\_lock);} \\
		\texttt{~~~~\_retry = true;} \\
		\texttt{~~~~tie(\_root,inserted) = \_insert(\_root,n); } \\
		\texttt{~~~~\_retry = false;} \\
		\texttt{~~~~pthread\_mutex\_unlock(\&\_tree\_lock);} \\
		\texttt{\}} \\
		\end{tabular}
	%\end{center}
	%\caption{mario man had a very large house. or a castle.}
	\caption{Unmodified code from {\tt htmavl.cpp} with the previously-unknown bug found by Landslide.
		The transaction path fails to check {\tt \_retry},
		leading to data races and corruption just as in {\tt htm1}.
		}
	\label{fig:avlbug}
\end{figure}
	%\item Despite Quicksand's ability to optimize for finding bugs in overall smaller state spaces,
	%	this does not correlate with better CPU-time or even better wall-clock time.
	%	While possibly a consequence of the modest test size,
	%	this suggests future work should extend the Quicksand algorithm
	%	to prioritize by state space maximality as well as by state space estimate.
	%\item

@@

			& 3,1 & \ETAdag{8d 11h} & \ETAdag{411312}	\\
			%[JOB 8] progress: 103873/411312 brs (25.254021%), ETA 8d 11h 36m 0s (elapsed 59m 42s)
			%[JOB 8] TIMED OUT (25.254021%; ETA 8d 11h 36m 0s)
			\hline
			{\tt avl\_insert}
			& 2,1 & \cpu{1083.35}	& \ints{40062}	\\
			%& 2,2 & \multicolumn{2}{c}{\em (bug!)} \\
			\hline
			{\tt avl\_fixed} % TODO
			& 2,1 & 	& 	\\
			& 2,2 & 	& 	\\
			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{10d 16h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{877.44}& \ints{28635}	\\
			& 2,2 & \ETAdag{2d 6h} & \ETAdag{5925634} \\
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

@@

\begin{table}[t]
	\begin{center}
	\begin{tabular}{l|c|r}
		\bf Landslide mode & \cpu{\bf cpu (s)} & \ints{\bf total int's} \\
		\hline
		{\sf verif} ({\tt -M})		& \cpu{~~14.95 $\pm$ 00.17} &  \ints{403} \\
		{\sf reinstrument}		& \cpu{~~24.72 $\pm$ 00.10} &  \ints{403}     \\
		{\sf quicksand} (no {\tt -M})	& \cpu{120.53 $\pm$ 18.62}  & \ints{2128}     \\
	\end{tabular}
	\end{center}
	\caption{Performance variance on {\tt counter}(2,2) in various modes.
		Average of 10 samples, $\pm N$ is 1 stddev.
	}
	\label{tab:variants2}
\end{table}
